The illusory promise of the Aligned Rank Transform

### 2025-12-20 sandbox.qmd
added several result subsections

```{=html}
<style>
.math.inline .MathJax  {
  font-size: 105% !important;
}
</style>
```

>>>>

```{=html}
<style>
.math.inline .MathJax  {
  font-size: 105% !important;
}
</style>
```
### Type I error rates in ordinal scales {#ordinal}
We also evaluate Type I error rates for ordinal scales. We test again a $4 \times 3$ within-subjects design.

**Main effects**. @fig-ordinal-main present Type I errors for $X_2$'s main effect, as we vary $X_1$'s magnitude of main effect.

::: {#fig-ordinal-main}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
prefix <- "Type_I_ordinal"

distributions <- c("likert_5", "likert_5_flex", "likert_7", "likert_7_flex", "likert_11", "likert_11_flex")
dnames <- c("5 - equidistant", "5 - flexible", "7 - equidistant", "7 - flexible", "11 - equidistant", "11 - flexible")
df <- readly_data(prefix, alpha, 1, distributions, dnames)

plotly_error(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 60)
```
Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** as a function of the magnitude $a_1$ of the main effect of $X_1$
:::

Our results indicate that PAR, RNK, and INT consistently maintain error rates close to $5\%$ across all tested ordinal scales. In contrast, ART exhibits a significant inflation of error rates, although this issue becomes less severe when the number of levels increases. Notably, error rates are more pronounced for flexible thresholds and tend to increase with sample size.

**Interaction effects**. @fig-ordinal-interaction-1 displays error rates for the interaction $X_1 \times X_2$ with a single main effect applied to $X_1$. The observed patterns align with our previous findings; ART consistently inflates error rates.

::: {#fig-ordinal-interaction-1}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}

plotly_error(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX1", max = 50)
```
Type I error rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitude $a_1$ of the main effect of $X_1$ 
:::

**Analysis for large samples.** Let us further explore ART's error rates when sample sizes become larger. As shown in @fig-ordinal-large-samples, error rates explode even when all population effects are null. The patterns are similar for all three effects.

::: {#fig-ordinal-large-samples}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
prefix <- "Type_I_4x3_ordinal_large_samples"

df <- readly_data_1(prefix, alpha, 0, distributions, dnames)
plotly_error_by_effect_2(df, xlab = "n", max = 100)
```
Type I error rates ($\alpha = .05$) as a function of the size $n$ of the sample, where all population effects are null.
:::

### 2025-12-20 text/50-ExperimentalMethod.qmd
added several result subsections

**Thresholds of ordinal scales.** For ordinal data, we evaluate $D =$ 5, 7, or 11 discrete levels. To derive the discretization thresholds $\tau_d$, we first consider the range $[-2SD, 2SD]$ , where $SD$ is the overall standard deviation of the latent variable $Y^{*}_{ijk}$. We then divide this range into 5, 7, or 11 intervals, following two different strategies: (i) setting thresholds to be *equidistant*; or (ii) considering *flexible* thresholds, randomly drawing their position in the above range.

@@

We do not directly evaluate Type II errors. Instead, we report on statistical *power* defined as $Power = 1 - \beta$, where $\beta$ is the rate of Type II errors. Significance tests do not provide any power guarantees. However, we can compare the power of different methods to evaluate their relative performance. In addition to power, we assess effect size estimates, where we use as ground truth the estimates of a parametric ANOVA conducted on the latent variable $\mathcal{Y}$. While partial $\eta^2$ is the most commonly used effect size measure, we also evaluate Cohen's $f$, as its expected value is proportional to the real magnitude of effect. However, $\eta^2$ can be directly derived from Cohen's $f$ as follows:

@@

As explained earlier, interaction effects are distorted when the latent variable $\mathcal{Y}$ is transformed to produce the observed responses. These transformations also influence the Type I error rates and statistical power we observe. In this analysis, we focus on evaluating these measures with respect to the effects applied to the latent variable $\mathcal{Y}$. We explicitly make this distinction whenever it is relevant to our discussion.

>>>>

**Thresholds for ordinal scales.** For ordinal data, we evaluate $D =$ 5, 7, or 11 discrete levels. To derive the discretization thresholds $\tau_d$, we first consider the range $[-2SD, 2SD]$ , where $SD$ is the overall standard deviation of the latent variable $Y^{*}_{ijk}$. We then divide this range into 5, 7, or 11 intervals, following two different strategies: (i) setting thresholds to be *equidistant*; or (ii) considering *flexible* thresholds, randomly drawing their position in the above range.

@@

We do not directly evaluate Type II errors. Instead, we report on statistical *power* defined as $Power = 1 - \beta$, where $\beta$ is the rate of Type II errors. Significance tests do not provide any power guarantees. However, we can compare the power of different methods to evaluate their relative performance. For log-normal and ordinal distributions, we also assess effect size estimates, where we use as ground truth the estimates of a parametric ANOVA conducted either on the logarithmic scale of the data (for log-normal) or latent variable $Y^*$ (for ordinal). While partial $\eta^2$ is the most commonly used effect size measure, we also evaluate Cohen's $f$, as its expected value is proportional to the real magnitude of effect. However, $\eta^2$ can be directly derived from Cohen's $f$ as follows:

@@

### 2025-12-20 text/60-Results-RatioScales.qmd
added several result subsections

### Type I error rates in ratio scales {#ratio}
We first evaluate Type I error rates for ratio scales. We test a 4 $\times$ 3 repeated-measures design, where we refer to the 4-level factor as $X_1$ and the 3-level factor as $X_2$.

@@

 The results show a very good behavior for RNK and INT across all distributions. The regular parametric ANOVA (PAR) keeps error rates below $5\%$. However, error rates become extremely low for some distributions, suggesting a loss in statistical power. We confirm previous results that ART fails to control the Type I error rate under the Cauchy distribution [@Mansouri:1995; @elkin:2021] but we also show that the method is problematic with other non-normal distributions. As the main effect on the first factor $X_1$ increases, Type I errors on the second factor $X_2$ grow and reach high levels. This pattern is particularly pronounced under the log-normal distribution. We also observe that for the binomial distribution, error rates are high ($\approx 11\%$ for $n = 20$) even when effects on $X_1$ are zero. In addition, error rates further grow when the sample size increases.

**Contrasts**. The same problems appear when we run ART's procedure for contrasts [@elkin:2021]. @fig-ratio-contrasts shows our results, where we report average error rates for three pairwise comparisons (since $X_2$ has three levels). In the rest of the paper, we only show results for overall effects, since results for contrasts exhibit the same patterns.

@@

**Interaction effects**. @fig-ratio-interaction-1 presents Type I error rates for the interaction effect $X_1 \times X_2$, when the main effect on $X_2$ is zero while the main effect on $X_1$ increases. Overall, we observe the same trends as for main effects. Again, ART fails in similar ways, although its error rates are now slightly lower.

@@

The scenarios above do not capture the failure cases of the rank transformation discussed earlier (see @fig-interactions-rank), as those arise only when both coefficients $a_1$ and $a_2$ are non-zero. Since there can be different possible definitions of an interaction effect in these cases, we address them in a separate subsection.

**Analysis for large samples.** Let us further explore ART's error rates when sample sizes become larger.

>>>>

### Type I error rates in ratio scales {#ratio}
We first evaluate Type I error rates for ratio scales. We test a 4 $\times$ 3 within-subjects design, where we refer to the 4-level factor as $X_1$ and the 3-level factor as $X_2$. We encourage readers to use the interactive tools provided in the figures to explore the trends in greater detail.

@@

Overall, RNK and INT exhibit very good behavior across all distributions. The standard parametric ANOVA (PAR) maintains error rates below $5\%$, but for some distributions (log-normal, Cauchy, and Poisson) the error rates are noticeably lower, suggesting a loss of statistical power. Consistent with previous findings [@Mansouri:1995; @elkin:2021], ART fails to control the Type I error rate under the Cauchy distribution. However, we further show that ART is also problematic for other non-normal distributions. As the main effect of the first factor $X_1$ increases, Type I errors on the second factor $X_2$ increase dramatically, approaching $100\%$ when $a_1=8$. For the binomial distribution, error rates are already elevated ($\approx 9\%$ for $n = 20$) even when $a_1=0$, and increase further as the sample size grows.

**Contrasts**. The same issues arise when running ART's contrast procedure [@elkin:2021]. @fig-ratio-contrasts shows the corresponding results, where we report average error rates for three pairwise comparisons (since $X_2$ has three levels). In the rest of the paper, we focus on overall effects, since the results for contrasts exhibit the same patterns.

@@

**Interaction effects**. @fig-ratio-interaction-1 presents Type I error rates for the $X_1 \times X_2$ interaction when $a_2=0$ and $a_1$ increases. Overall, we observe trends similar to those for main effects. ART again fails in comparable ways, although its error rates are slightly lower in this case. In addition, PAR tends to inflate error rates under the log-normal and exponential distributions as $a_1$ increases.

@@

The scenarios above do not capture the failure cases of the rank transformation discussed earlier (see @fig-interactions-rank), as these arise only when both coefficients $a_1$ and $a_2$ are non-zero. Because multiple definitions of interaction effects are possible in such cases, we address them in a separate subsection.

**Analysis for large samples.** We also examine Type I error rates for larger sample sizes (up to $n=512$), fixing $a_1=1$. As shown in @fig-ratio-large-samples, ART's error rates increase only slightly for log-normal data but rise substantially under the Cauchy, Poisson, and binomial distributions. These patters are consistent for both the main effect of $X_2$ and the interaction effect. As we later observe for ordinal scales, ART's tendency for increasing Type I error rates with sample size appears to be a more general issue for discrete distributions.

### 2025-12-20 text/62-Results-AcrossExpDesigns.qmd
added several result subsections

### Type I errors across experimental designs {#extra}
We investigate the generalizability of the aforementioned results to other experimental designs involving two or three factors. We assess five out of the six ratio scales examined earlier. We exclude the Cauchy distribution and replace it by a 5-level ordinal scale with flexible thresholds.

**Main effects**. @fig-designs-main illustrates Type I errors for the main effect of $X_2$ while varying the magnitude of the main effect of $X_1$, with a focus on $n = 20$. Across all cases, RNK and INT consistently maintain error rates close to $5\%$. PAR's error rate remains overall close to $5\%$ but reaches notably low levels for specific combinations of designs ($2 \times 2 \times 2$ and $3 \times 3 \times 3$ repeated-measures) and distributions (log-normal and exponential). In contrast, ART inflates error rates for all non-normal distributions, with discrepancies across different designs. We observe that ART is particularly problematic in discrete distributions when applied to the $2 \times 2 \times 2$ and $3 \times 3 \times 3$ designs.

>>>>

### Type I errors across experimental designs {#extra}
We examine the generalizability of the preceding results across alternative experimental designs involving two or three factors. We assess five of the six ratio-scale distributions considered earlier, excluding the Cauchy distribution and replacing it with a 5-level ordinal scale with flexible thresholds.

**Main effects**. @fig-designs-main illustrates Type I error rates for the main effect of $X_2$ while varying the magnitude of the main effect of $X_1$, focusing on $n = 20$. RNK and INT consistently maintain error rates close to $5\%$, with the exception of the Poisson distribution, for which error rates become low in the two within-subjects designs (particularly for INT). Similarly, PAR generally maintains error rates near $5\%$, but they drop to notably low levels for certain combinations of designs ($2 \times 2 \times 2$ and $3 \times 3 \times 3$ within-subjects) and distributions (log-normal and exponential).

In contrast, ART inflates error rates across all non-normal distributions, with the extent of inflation varying by design. For example, ART performs particularly poorly for the ordinal scale when applied to the $2 \times 2 \times 2$ and $3 \times 3 \times 3$ designs. In these designs, we also observe inflated error rates for log-normal data, even when $a_1 = 0$.

### 2025-12-20 text/66-Results-Interactions.qmd
added several result subsections

### Interactions under parallel main effects {#interactions}
Let us now examine interaction effects when both coefficients $a_1$ and $a_2$ increase in parallel. In this scenario, the interpretation of an interaction effect can become ambiguous. Here, we define the null hypothesis in terms of the coefficients of the latent linear model, following the recommendation of @Greene:2010 --- that is, $a_{12} = 0$.

@@

Error rates become exceptionally high in some cases, reaching up to 100%. However, these results require careful interpretation. As discussed in [Section 4](#interpretation), interaction effects in nonlinear models can be interpreted in two distinct ways, and the results from the different methods are not consistent with our definition. Let us examine the outcomes for each method:

- *PAR*. Error rates are high for all non-normal distributions and become even higher when the sample size increases. However, this behavior clearly stems from the method's sensitivity to the scale of the observations, rather than to the model parameters.

- *RNK.* Error rates explode when both main effects exceed a certain level (e.g., when $a_1, a_2 \ge 4$ and $n = 20$). This problem arises from the way the rank transformation distorts interaction effects (see @fig-interactions-rank). For all continuous distributions, we observe that error rates exhibit the exact same patterns.

- *INT.* It performs better than RNK. Errors begin to increase across all distributions only when the main effects become large. An exception is the binomial distribution, for which the error rates of INT and RNK are similar.

- *ART.* It maintains correct error rates as long as population distributions are normal. For all other distributions, error rates increase rapidly as effect sizes grow. Two factors explain this behavior: (i) a lack of statistical robustness to violation of the method's assumptions, as also observed in our previous experiments; and (ii) interaction interpretation issues. These results confirm that ART is not scale-invariant, interpreting interactions relative to the scale of the observed data.

@@

**Other designs.** Our findings in @fig-designs-interaction-2 further demonstrate that all methods struggle to accurately detect interactions for different reasons. Interestingly, INT shows unusually low error rates in certain configurations (e.g., $2 \times 3$ and $2 \times 2 \times 2$), indicating potential difficulty in detecting subtle interaction effects when main effects are large. RNK follows a similar trend for the $2 \times 2 \times 2$ design. Both RNK and INT also exhibit higher error rates under ordinal and binomial scales, though these remain lower than those observed for PAR and ART.

@@

Some readers may argue that evaluating PAR and ART using an interaction definition that accounts for the scale of the observed variable might be more appropriate. We offer three responses to this criticism:

1. It is unclear how to reliably control the null hypothesis at the level of the response scale when both main effects are varied.

2. Our earlier experiments demonstrate that ART fails to maintain nominal Type I error rates for interactions even when only a single main effect is present. Introducing a second main effect could only exacerbate the problem.

3. The purpose of these results is not to provide precise benchmarks, but rather to illustrate the complexity and pitfalls of interpreting interaction effects under assumption violations --- regardless of the method used. Many researchers are unaware of these issues and often misinterpret the $p$-value when testing interactions. Note that @elkin:2021 compared ART to ANOVA on log-transformed data under the belief that ART tests null hypotheses about parameters of the latent linear model. Our findings clearly show that this is not the case.

>>>>

### Interactions under parallel main effects {#interactions}
Let us now examine interaction effects when both coefficients $a_1$ and $a_2$ increase in parallel. In this scenario, the interpretation of an interaction effect becomes ambiguous in nonlinear models. Here, we define the null hypothesis in terms of the coefficients of the linear component of the model (see @eq-linear-model), following the recommendation of @Greene:2010 --- that is, $a_{12} = 0$.

@@

Error rates become exceptionally high in many cases, reaching up to 100%. These results require careful interpretation. As discussed in [Section 4](#interpretation), interaction effects in nonlinear models can be interpreted in two distinct ways, and the results produced by the different methods are not alwyas consistent with our chosen definition. We therefore examine the outcomes for each method in turn:

- *PAR*. Error rates are high for all non-normal distributions (with the exception of the Cauchy distribution) and increase further as sample size grows. This behavior arises because the null hypothesis for this method is defined on the scale of the response variable, rather than on the scale of the linear model.

- *RNK.* Error rates increase dramatically once main effects exceed a certain magnitude. This well-known issue stems from the way rank transformations distort interaction effects (see @fig-interactions-rank), even when the underlying distributions are normal.

- *INT.* It performs better than RNK in continuous distributions, with errors rates increasing only when $a_1$ and $a_2$ become large. However, under the Poisson and binomial distributions, INT performs worse than RNK. Interestingly, INT is particularly robust under the Cauchy distribution.

- *ART.* ART maintains correct error rates only when the population distributions are normal. For all other distributions, error rates increase rapidly as effect sizes grow. Two factors may account for this behavior: (i) a lack of robustness to violations of the method’s assumptions, as also observed in our earlier experiments; and (ii) issues related to interaction interpretation. These findings confirm that ART is not scale-invariant and interprets interactions relative to the scale of the observed data.

@@

**Other designs.** Our findings in @fig-designs-interaction-2 further demonstrate that all methods struggle to accurately detect interactions. INT shows unusually low error rates in certain configurations (e.g., $2 \times 3$ and $2 \times 2 \times 2$), suggesting potential difficulty in detecting subtle interaction effects when main effects are large. RNK displays a similar pattern for the $2 \times 2 \times 2$ design. In addition, both RNK and INT exhibit higher error rates under discrete distributions, in some cases exceeding those observed for ART.

@@

These results illustrate the complexity and pitfalls of interpreting interaction effects under assumption violations --- regardless of the method used.
